---
title: "Making a github webpage with Rstudio and Hugo"
author: "Hicham Zmarrou"
date: "March 25, 2017"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

```{r, include = FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

-   [Introduction](#introduction)
    -   [Blogdown](#blogdown)
    -   [Hugo](#hugo)
    -   [GitHub](#github)
-   [Building the Page](#building-the-page)
    -   [Installing Necessary Packages and Software](#installing-necessary-packages-and-software)
        -   [Setting Up GitHub](#setting-up-github)
        -   [RStudio and Blogdown](#rstudio-and-blogdown)
    -   [Initializing GitHub pages to work with Hugo and Blogdown](#initializing-github-pages-to-work-with-hugo-and-blogdown)
-   [Adding Content](#adding-content)
    -   [Site Configuration](#site-configuration)
    -   [Adding a Page](#adding-a-page)
    -   [Page Templates and Fixing Blank Pages](#page-templates-and-fixing-blank-pages)
    -   [Adding a Blog Post or Portfolio Piece](#adding-a-blog-post-or-portfolio-piece)
    -   [Adding images](#adding-images)
-   [Troubleshooting](#troubleshooting)
    -   [Page is Blank](#page-is-blank)
    -   [Images Aren't Showing Up](#images-arent-showing-up)
    -   [Page Has No Formatting](#page-has-no-formatting)
    -   [My Post Won't Show Up](#my-post-wont-show-up)
    

## Introduction

As I transitioned into the data science world, I kept hearing one piece of advice over and over: build a portfolio. So, once I had completed a few personal projects, I decided to investigate the best way to build a portfolio that could also house blog posts, comments, and (naturally) display R code and outputs seamlessly.

I looked into [Wordpress](https://wordpress.com/) but I wasn't thrilled with the theme choices available if I opted for free site hosting.  Then I realized that you can generate your site entirely using [RStudio](http://rmarkdown.rstudio.com/rmarkdown_websites.html), which seemed like a great option but had limited customization options. 

Then I stumbled upon [Blogdown](https://github.com/rstudio/blogdown), a new RStudio package developed by `knitr` and `bookdown` creator [Yihui Xie](https://github.com/yihui).  The package runs using a static-site generator called "[Hugo](https://gohugo.io/)" which, when I started making my site, I had never heard of. And while the `blogdown` package was still under development and had no formal package documentation, I wanted to give it a try.  After several days of googling, experimenting, and learning, I am now the proud owner of a site I love.  This post details how I made it happen. 

<i class="fa fa-arrow-circle-right" aria-hidden="true"></i> ***Disclaimer: My method may not be the most stream-lined at this time, and I fully expect to look back at this post in the future and wonder how I ever got anything done, but for the time being, this method works.*** 

Alright, before I delve into the *how* of it all, I wanted to give a little bit of background information on the pieces that are involved in this: GitHub, Hugo, and blogdown. If you want to cut right to the chase, go [ahead](#building-the-page). 


### GitHub

First of all, I do want to provide some background information about what Github does and why it's useful. Skip this section if  you're a GitHub pro. 

First, GitHub is powered by Git, a command-line version-control system.  If you're not familiar with it, Git has no buttons to press, only commands to type to control it, but it allows you to save multiple versions of a file.  The beauty of version control is that if you mess up, you can go back to the last saved version.  It's like reaching a "save point" in a video game that you can return to if your character dies. GitHub is the place where you can store all of these versions of your work. Other developers can see, make a copy of, ammend their copy, and make suggestions on your work.

<i class="fa fa-arrow-circle-right" aria-hidden="true"></i> **Keep in mind**: GitHub does have a free version, but it makes all of your work publicly available. For instance, you can go dig around in all of the code fot this course [here](https://github.com/trefoil-ml). 

Ok, [basic vocab](https://help.github.com/articles/github-glossary/) for GitHub use and then I'll promise I'll get started!

* **Repository** : When you make an account on [Github](https://github.com/) the first thing you'll want to do is make a repository (or repo). This is the place to store all of your files for a given project.
* **Forking** : Say you really like someone else's project on GitHub and you'd like to make a copy of it on your account to adjust however you'd like? Well, that process is called forking. To do it, find a repo that you like and click the ' <i class="fa fa-code-fork" aria-hidden="true"></i> fork ' button in the right hand corner.
* **Branch** : This is a "parallel version" of a repo that you can adjust without impacting the original repo. 
* **Remote** : The copy of your files that reside on GitHub.com
* **Local** : The copy of your files that reside on your computer
* **Commit** : A change to a file, usually submitted with a message from you to indicate what was changed
* **Push** : Once you commit changes on your local files, you want to send (or push) them to your remote repo, making them available for others.
* **Pull** : If multiple people are working in your remote repo, they may have made a change that is not reflected in your local version. You can pull the newest version down to your computer to work on it. 
* **Subtree** : This is a repo inside of a repo. More on this later. 

Ok, all of that will be helpful as we continue! 

## Building the Page

### Installing Necessary Packages and Software

